Results section

### 2022-10-18 code/Matrinem data analysis.Rmd
Created new result script

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading libraries}
library(readxl)
library(tidyverse)
library(broom)
library(ggpubr)
#library(AICcmodavg)
library(visdat)
library(rstatix)
library(gt)
library(gtsummary)
library(patchwork)
```

# Introduction

|           | PreExp 1 | PreExp 2 | PreExp 3 | NEC-FvT |
|-----------|----------|----------|----------|---------|
| **Aim**   |          |          |          |         |
| Protocols |          |          |          |         |
| N         |          |          |          |         |

## Pre-Experiment 1 (PreExp 1)

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r loading libraries}
library(readxl)
library(tidyverse)
library(broom)
library(ggpubr)
#library(AICcmodavg)
library(visdat)
library(rstatix)
library(gt)
library(gtsummary)
library(patchwork)
library(jtools)
```

# MATRINEM - Attempts at creating a NEC mouse model

## Preface

## Introduction

This thesis is about...

## Background

## Methods and Materials

Prior to results are presented a short description of the studies conducted needs to be made.

Experiment no. 4 was the goal, but before we could test our overall hypothesis we needed to find out which combination of antibiotics and feeding scheme would results in the most NEC-like phenotype.

These research questions were investigated in 3 pre-experiments (PreExp 1-3), each with their own research questions.

### **Overview of all experiments**

+-----------+-------------------------------------------------------------------------------+----------+----------+---------+
|           | PreExp 1                                                                      | PreExp 2 | PreExp 3 | NEC-FvT |
+===========+===============================================================================+==========+==========+=========+
| **Aim**   | To test feasibility of formula feeding with PICC line                         |          |          |         |
+-----------+-------------------------------------------------------------------------------+----------+----------+---------+
| Protocols | Testing a broad spectrum antibiotics mix in combination with formula feeding. |          |          |         |
|           |                                                                               |          |          |         |
|           | 4 groups                                                                      |          |          |         |
|           |                                                                               |          |          |         |
|           | 1.  CON-BF                                                                    |          |          |         |
|           | 2.  CON-FF                                                                    |          |          |         |
|           | 3.  AB-BF                                                                     |          |          |         |
|           | 4.  AB-FF                                                                     |          |          |         |
+-----------+-------------------------------------------------------------------------------+----------+----------+---------+
| N         | 38 (32 survived)                                                              |          |          |         |
+-----------+-------------------------------------------------------------------------------+----------+----------+---------+

### PreExp 1 - Feasibility Study

#### **Rationale:**

*Chen et al.*, showed that Maternal Antibiotics Treatment (**MAT**) resulted in mild NEC-like symptoms of mice offspring. We believed that MAT combined with formula feeding (**FF**) of offspring would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}

#### **Aim:**

To test if it was possible to keep mice separated from the mother alive from postnatal day 3 and 48 hours onwards.

#### **Research Questions:**

-   Can we keep offspring alive with formula feeding?
-   Will combining MAT with FF result in a more inflammatory intestine as measured by cytokine analysis in the ileum.

@@

**Results**: `r inline_text(tbl1, {N})` mice were born to 6 different mothers. Of these 32 survived and were included in the analysis. Cytokines

```{r}
cytokines_1 <- read_excel(path = "data/processed/cytokines.xlsx") %>%  #Load in cytokine data for all experiments
  filter(exp_number == 1) %>% 
  filter(tissue == "ileum") %>% 
  mutate(
    group = recode_factor(  #Change the order of the groups to desired plot order
      group,
      "CON-BF"= "CON-BF",
      "CON-FF" = "CON-FF",
      "AB-BF"= "AB-BF",
      "AB-FF"= "AB-FF"
      ))

cytokines_1_long <- cytokines_1 %>% 
  pivot_longer(cols = INFg:TNFa, 
               names_to = "cytokines", 
               values_to = "cytokine_concentration") %>% 
  drop_na(cytokine_concentration)

normality_check <- cytokines_1_long %>% 
  group_by(cytokines) %>% 
  shapiro_test(cytokine_concentration) %>% 
  select(cytokines, p) %>% 
  filter(p < 0.05)

normality_plot <- cytokines_1_long %>%
  filter(cytokines %in% normality_check$cytokines ) %>% 
  ggplot(aes(x = group, 
             y = cytokine_concentration,
             color = group)) + 
  geom_jitter() +
  facet_wrap(~ cytokines,
             scales = "free") + 
  labs(title = "Non-normally distributed cytokines (Shapiro-Wilk Normality Test, p < 0.05)",
       y = "Concentration of cytokine in tissue (pg/mL)",
       x = "") +
  theme_bw()

normality_plot

normality_result_1 <- normality_check %>% 
  knitr::kable(caption = "Non-normally distributed cytokines (Shapiro-Wilk Normality Test, p < 0.05)")
normality_result_1
```

```{r}
tbl1 <- cytokines_1 %>% 
  select(group,INFg:TNFa) %>% 
  tbl_summary(by = group,
              digits = everything() ~ c(2,1,1),
              missing = "no",
              include = !"INFg") %>%
  add_p() %>% 
  modify_header(
    label ~ "Cytokines") %>% 
  modify_spanning_header((c("stat_1", "stat_2", "stat_3", "stat_4") ~ "Total number of mice analyzed = {N}"))

  tbl1_gt <- tbl1 %>% 
  as_gt() %>%
  tab_header(title = md("**Proinflammatory cytokines**")) %>%
  opt_align_table_header("center") %>% 
  tab_options(heading.title.font.size = 20,
              heading.background.color = "lightgrey",
              ) %>% 
  tab_source_note(
    source_note = " * P-value indicates overall group differences"
  ) %>%
  tab_source_note(
    source_note = md("**Matrinem round 1**")
  ) %>%
  cols_width(label ~ px(100),
             stat_1 ~ px(150),
             stat_2 ~ px(150),
             stat_3 ~ px(150),
             stat_4 ~ px(150)) 
tbl1_gt
# Save and export files
# tbl1 %>%  #Save as excel file.
#   as_hux_table() %>%
#   huxtable::quick_xlsx(., file = "../Results/Round 1/ileum-cytokines.xlsx")
# 
# tbl1 %>% #Save as image file.
# as_flex_table() %>%
#   flextable::save_as_image(path = "../../Results/Round 1/ileum_cytokines_table.png",webshot = "webshot")
```

```{r plots all cytokines, warning=FALSE}
plot_all_cytokines <- cytokines_1_long %>%
  #filter(!variables %in% "INFg") %>%  #Exclude any cytokines not wanted in the plot
ggplot(aes(x = group, 
           y = cytokine_concentration, 
           fill = group)) + 
  geom_boxplot() +
  facet_wrap(~cytokines, 
             scales = 'free', 
             nrow = 3,
             ncol = 4) +
  labs(title = "Levels of pro-inflammatory cytokines in the ileum",  # Change title, and axis names
       x = "",
       y = "Tissue conc. (pg/mL)",
       fill = "Groups") +
  theme_bw() +
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    axis.text.y = element_blank(),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="white")) + # Change legend box background color
  scale_fill_brewer(palette="Set1")

all_cytokines_patch <- plot_all_cytokines + plot_annotation (
  title = 'Cytokine expression',
   subtitle = 'All cytokines measured',
  tag_levels = 'A',
  caption = 'CON = Water, AB = Antibiotics, BF = Breast-Feeding, FF = Formula-Feeding',
  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))

all_cytokines_patch
#ggsave('../../Results/Round 1/All Cytokines-boxplot.png',all_cytokines_patch)
```

```{r}
#tbl1_p contains cytokines with significant changes
tbl1_p <- tbl1 %>% 
  filter_p() %>% 
  as_tibble()

signif_cytokines <- cytokines_1_long %>%
  filter(cytokines %in% tbl1_p$Cytokines) %>%  #Exclude any cytokines not wanted in the plot
ggplot(aes(x = group, 
           y = cytokine_concentration, 
           fill = group)) + 
  geom_boxplot() +
  stat_compare_means(aes(label = ..p.signif..),
                   method = "wilcox.test", ref.group = "CON-BF",
                   label.y.npc = 0.55) +
  stat_compare_means(
    aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
    label.y.npc = 0.9,
    label.x.npc = 0.5) +
  facet_wrap(~cytokines, 
             scales = 'free', 
             nrow = 3,
             ncol = 3) +
  labs(title = "Levels of pro-inflammatory cytokines in the ileum", # Change title, and axis names
       #subtitle = "Statistically significant changes", 
       x = "",
       y = "Tissue conc. (pg/mg)",
       fill = "Groups") +
  theme_bw()+
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    axis.text.y = element_blank(),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="white")) + # Change legend box background color
  scale_fill_brewer(palette="Set1")

signif_cytokines_patch <- signif_cytokines + plot_annotation (
  title = 'Cytokine expression',
  subtitle = 'Statistically significant changes',
  tag_levels = 'A',
  caption = 'CON = Water, AB = Antibiotics, BF = Breast-Feeding, FF = Formula-Feeding. \n
  Pairwise comparisons with CON-BF using wilcoxon test',
  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))
signif_cytokines_patch
#ggsave('../../Results/Round 1/Significant Cytokines-boxplot.png',signif_cytokines_patch)
```

```{r echo=FALSE, warning=FALSE}
#### Make regression model on long format
reg_mod_cyt_1 <- cytokines_1_long %>% 
  group_by(cytokines) %>% 
  do(lm(data = ., cytokine_concentration ~ maternal_treatment*type_of_feed) %>% tidy)  %>% filter(term!='(Intercept)') %>% 
  ungroup() %>% 
  filter(p.value < 0.05)

reg_mod_cyt_1 %>% 
  select(-statistic) %>% 
  knitr::kable(., caption = 'Significant effectors on cytokine expression', digits = 3)
```

Statistically significant group differences were observed for IL-6 (`r inline_text(tbl1, column = "p.value", variable = "IL_6")`), KCGRO (`r inline_text(tbl1, column = "p.value", variable = "KCGRO")`) and TNFa (`r inline_text(tbl1, column = "p.value", variable = "TNFa")`).

**The conclusion** is that many of the cytokines show an interaction effect of the two treatment parameters

#### Summary of PreExp 1

-   Can we keep offspring alive with formula feeding?

    -   After fine-tuning the feeding method we we able to successfully feed all surviving (32) mice every 3 hours.
    -   The formulafed animals gained weight - albeit less than their breastfed peers - during the 48 hours.
    -   MAT did not affect growth. CON-BF and AB-BF gained weight at a similar pace.

-

    ## Will combining MAT with FF result in a more inflammatory intestine as measured by cytokine analysis in the ileum.

### PreExp 2 - Selectively targeting gram+ or gram- bacteria
